Unshuffling Permutations

### 2015-06-26 article.tex
figure

\usepackage[utf8x]{inputenc}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
%\usepackage{llncsdoc}
\usepackage[english]{babel}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{algorithmic}
\usepackage{float}
\usepackage{todonotes}
\usepackage{subfigure}
%\usepackage[mathcal]{euscript}
\usepackage{color}
\newcommand{\rt}{\textcolor{red}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{matrix}
\input{math}
\input{complexity}

@@

\newcommand{\PET}{pet}
\newcommand{\SBL}{sbl}
\newcommand{\horiz}{{\centering\rule{350pt}{.4pt}}}

@@

\title{%
Obtaining a Triangular Matrix by Independent Row-Column Permutations
}%

\author{%
  Guillaume Fertin\inst{1} \and
  Irena Rusu\inst{1} \and
  St\'ephane Vialette\inst{2}
}% end author
\institute{%
  LINA UMR CNRS 6241, Universit\'e de Nantes, France \\
  \email{\{guillaume.fertin,irena.rusu\}@univ-nantes.fr}
  \and
  Universit\'e Paris-Est, LIGM (UMR 8049), CNRS, UPEM, ESIEE Paris, ENPC, 
  F-77454, Marne-la-Vallée, France\\
  \email{vialette@univ-mlv.fr}
}% end institute
\date{\today}

@@

\begin{abstract}
Given a square $(0,1)$-matrix $A$,
we consider the problem of deciding whether there exists a permutation
of the rows and a permutation of the columns of 
$A$ such that after carrying out these permutations,
the resulting matrix is triangular. The complexity of the problem was
posed as an open question by Wilf~\cite{Wilf:1997} in 1997.
%(\emph{i.e.}, do there exist permutation matrices $P$ and $Q$ such that
%$PAQ$ is triangular)?
In 1998, DasGupta et al.~\cite{DasGuptaJiangKannanLiSweedyk:1998} seemingly answered the
question, proving it is \NPC. However, we show here that their result is flawed, which 
leaves the question still open. Therefore, we give a definite answer
to this question by proving that the problem is \NPC. We finally present an
exponential-time algorithm for solving the problem.
\end{abstract}

@@

In his contribution in the tribute to the late Professor
Erd\"os~\cite{Wilf:1997}, Wilf posed the following question:
\emph{
  ``Let $A$ be a $m \times n$ matrix of $0$'s and $1$'s. Consider the
  computational problem: do there exist permutations $P$ of the rows
  of $A$, and $Q$, of the columns of $A$ such that after carrying out
  these permutations, $A$ is triangular? The question we ask concerns
  the complexity of the problem.
  Is this problem \NPC?
  Or, does there exist a polynomial-time algorithm for doing it?''}
As noted by Wilf,
this problem is strongly related to job scheduling with precedence
constraints, a well-known problem in theoretical computer science.
The present paper is devoted to giving an answer to this question.

A square matrix is called {\em lower triangular} if all the entries above the main diagonal are zero.
Similarly, a square matrix is called {\em upper triangular} if all the entries below the main diagonal
are zero.
A {\em triangular matrix} is one that is either lower triangular or upper triangular.
Because matrix equations with triangular matrices are easier to solve,
they are very important in  linear algebra and numerical analysis.
We refer the reader to \cite{MatrixComputations:1996} for a further discussion.

For an arbitrary square matrix $A$, it is well-known in linear algebra
that there exists an invertible matrix $S$ such that $S^{-1}AS$ is
upper triangular.
We focus here, however, on permutation matrices.
Recall that a permutation matrix is a square matrix obtained from the same
size identity matrix by a permutation of rows. 
A product of permutation matrices is again a permutation matrix and
the inverse of a permutation matrix is again a permutation matrix.
In fact, for any permutation matrix $P$, $P^{-1} = P^T$.

% Nicolson [5] shows that for a nonnegative matrix $A$ the following four conditions
% are equivalent:
% (i) there exists a permutation matrix $P$ such that
% $P A P^T$ is strictly upper triangular;
% (ii) there is no positive cycle in $M$
% (i.e. in the weighted digraph there is no cycle whose edges all have positive weight);
% (iii) the permanent of $M + I$ is $1$; and
% (iv) $A$ is nilpotent.

This paper is organized as follows.
In Section~\ref{section:Notations}, we provide the basic material
needed for this paper. %with a
%strong emphasis on the associated combinatorial structures: configurations of subsets,
%bipartite graphs and directed graphs.
Section~\ref{section:Hardness result} is devoted to proving hardness of
determining whether a square $(0,1)$-matrix is permutation equivalent
triangular, i.e. whether it can be transformed into a triangular matrix by independent
row and column permutations.
In Section~\ref{section:Exponential-time algorithm}, we give some properties of
permutation equivalent triangular matrices (or \PET\ matrices, for short) and present 
an exponential-time algorithm to determine whether a matrix is a \PET\ matrix.
The paper concludes with suggestions for further research directions.

@@

%%
%% ---- Preliminaries ----
%%
\section{Notations}
\label{section:Notations}

%\subsection{$(0,1)$-matrices}
%\subsubsection{$(0,1)$-matrices} 
For any positive integer $n$, denote $[n]=\{1, 2, \ldots, n\}$. 
Let $A = [a_{i,j}]$, $1 \leq i \leq m$ and $1 \leq j \leq n$,
be a matrix of $m$ rows and $n$ columns.
% We say that $A$ is of size $m$ by $n$ and we also refer to $A$
% as an $m$ by $n$ matrix.
In the case that $m = n$ then the matrix is \emph{square} of \emph{order} $n$.
It is always assumed that the entries of the matrix are elements
of some underlying field $F$.
It is convenient to refer to either a row or a column of the matrix as a
\emph{line} of the matrix.
We use the notation $A^T$ for the \emph{transpose} of matrix $A$.
We always designate a zero matrix by $\zeros$, a matrix with every entry equal to
$1$ by $J$, and the identity matrix of order $n$ by $I$.
In order to emphasize the size of these matrices we sometimes include subscripts.
Thus $J_{m,n}$ denotes the all $1$'s matrix of size $m$ by $n$,
and this is abbreviated to $J_n$ if $m = n$. 
Notations $\zeros_{m,n}$, $\zeros_n$ and $I_n$ are similarly defined.
In displaying a matrix we often use $*$ to designate a submatrix of no particular structure.
Two matrices $A$ and $B$ are said to be \emph{permutation equivalent},
denoted by $A \sim B$, if there exist permutation matrices $P$ and $Q$ of
suitable sizes such that $B = PAQ$.

We will be greatly concerned with matrices whose entries consist exclusively of
the integers $0$ and $1$.
Such matrices are referred to as \emph{$(0,1)$-matrices}.
For a $(0,1)$-matrix $A$, we let $\omega(A)$ stand for the number of $1$'s in $A$.
A square matrix $A = [a_{i,j}]$ of order $n$ is said to be
\emph{lower left triangular} if it has only $0$'s above the main diagonal
(\emph{i.e.} $a_{i,j} = 0$ for $1 \leq i < j \leq n$.
We write $\lltm_{\,n}$ for the lower left triangular $(0,1)$-matrix 
whose $0$'s are exclusively above the main diagonal.
For two matrices $A = [a_{i,j}]$ and $B = [b_{i,j}]$ of size $m$ by $n$,
we write $A \leq B$ if $a_{i,j} \leq b_{i,j}$ for $1 \leq i \leq m$
and $1 \leq j \leq n$, so that a square matrix $A$ of order $n$ is
lower left triangular if $A \leq \lltm_{\,n}$.
In the context of permutation equivalent matrices,
we will sometimes not be interested in any particular orientation of a
triangular matrix and forget about any
specific orientation such as ``\emph{lower left}''.
Furthermore, for readability, a matrix which is permutation equivalent
to a triangular matrix is said to be a \emph{\PET\ matrix}.
The \emph{row sum vector}
$\RSV(A) = \begin{bmatrix} r_1 & r_2 & \ldots & r_m \end{bmatrix}$
and the  \emph{column sum vector}
$\CSV(A) = \begin{bmatrix} c_1 & c_2 & \ldots & c_n \end{bmatrix}$
of $A$ are defined by
$r_i = \sum_{1 \leq j \leq n} a_{i,j}$ for $1 \leq i \leq m$
and
$c_j = \sum_{1 \leq i \leq m} a_{i,j}$ for $1 \leq j \leq n$.
The row sum vector $\RSV(A)$
(resp. column sum vector $\CSV(A)$)
is \emph{stepwise bounded}
if  $|\{i : r_i \leq k \}| \geq k$
(resp. $|\{j : c_j \leq k \}| \geq k$)
for $1 \leq k \leq n$.
It is clear that if a $(0,1)$-matrix $A$ is a \PET\ matrix then
both $\RSV(A)$ and $\CSV(A)$ are stepwise bounded.

% The definition of the \emph{permanent} of $A$ follows the classical definition of the
% determinant but with a simplification. Ironically, this simplification in the formula makes
% it more difficult to compute the permanent.
The permanent of $A = [a_{i,j}]$ is defined as the number given by the formula
$\PER(A) = \sum_{(j_1, j_2, \ldots, j_n) \in S_n}
a_{1,j_1}\,a_{2,j_2}\, \ldots a_{n,j_n}$,
where the summation is over all permutations
$(j_1,j_2, \ldots, j_n)$ of $[n]$.
Observe that, unlike the determinant, we do not put a minus sign in front of some of the terms in the
summation.
Of particular importance,
the permanent does not change if we permute the rows of $A$ and
permute the columns of $A$.

Let $A = [a_{i,j}]$ be an $m$ by $n$ matrix.
For convenience, for a set $K \subseteq [m]$ we will
write $\overline{K}$ for the set
$[m] \setminus K$.
Let $K = \{i_1, i_2, \ldots, i_k\}$ be a set of $k$ elements with
$K \subseteq [m]$,
and let $L = \{j_1, j_2, \ldots, j_l\}$ be a set of
$l$ elements with $L \subseteq [n]$.
The sets $K$ and $L$ designate a collection of row indices and column indices,
respectively, of the matrix $A$, and the $k$ by $l$ submatrix determined by them is
denoted $A[K,L]$.
% $$
% A[K,L]=
% \begin{bmatrix}
% a_{i_1,j_1} & a_{i_1,j_2} & \dots  & a_{i_1,j_l} \\
% a_{i_2,j_1} & a_{i_2,j_2} & \dots  & a_{i_2,j_l} \\
% \vdots      & \vdots      & \ddots & \vdots      \\
% a_{i_k,j_1} & a_{i_k,j_2} & \dots  & a_{i_k,j_l}
% \end{bmatrix}\text{.}
% $$
% If $L = K$, then $A[K, K]$ is a \emph{principal} submatrix of $A$,
% sometimes denoted more simply as $A[K]$.

%\subsection{Related combinatorial objects}
%\subsubsection{Configuration of subsets}

% Combinatorial matrix theory \cite{Brualdi:Ryser:1991} includes a wide variety of 
% combinatorial areas, e.g. graph theory \cite{Diestel:2000}, transversal theory
% \cite{Mirsky:1971}, combinatorial configurations \cite{Ryser:1969}.
% We shall make intensive use of configuration of subsets.

Let $X = \{x_i : 1 \leq i \leq n\}$ be a non-empty set of $n$
elements, that we call an \emph{$n$-set}.
Let $\mathcal{S} = (S_i : 1 \leq i \leq m)$ be $m$ not necessarily
distinct
subsets of the $n$-set $X$. We refer to this collection of subsets of an
$n$-set as a \emph{configuration of subsets}.
We set $a_{i,j} = 1$ if $x_j \in S_i$, and $a_{i,j} = 0$
if $x_i \notin S_i$.
The resulting $(0,1)$-matrix $A = [a_{i,j}]$, $1 \leq i \leq m$ and
$1 \leq j \leq n$ of size $m$ by $n$ is the \emph{incidence matrix}
for the configuration of subsets $\mathcal{S}$ of the $n$-set $X$.
The $1$s in row $\alpha_i$ of $A$ display the elements in the subset
$S_i$, and the $1$'s in column $\beta_j$ display the occurrences of $x_j$
among the subsets. 
Let $\mathcal{S} = (S_i : 1 \leq i \leq n)$ be a configuration of subsets of some 
ground $n$-set $X$. A bijective mapping $\varphi : \mathcal{S} \to [n]$ is said to be a
\emph{stepwise bounded labeling} (or \SBL\ for short) of $\mathcal{S}$ if
$\left|\bigcup_{\varphi(S_j) \leq i}S_j\right| \leq i$
for $1 \leq i \leq n$.

%  A \emph{system of distinct representatives} (SDR) for the family $\mathcal{S}$
% is a family $(x_1, x_2, \ldots, x_n)$ of elements of $X$ satisfying
% the conditions
% (i) $x_i \in S_i$ for $1 \leq i \leq n$, and
% (ii) $x_i \neq xj$ for $i \neq j$.
% (The first condition asserts that the elements are representatives of the sets,
% and the second that they are distinct.)
% The collection $\mathcal{S}$ satisfies the \emph{marriage} condition
% if and only if for each subcollection $\mathcal{S}' \subseteq \mathcal{S}$, we have
% $|\mathcal{S}'| \leq \left|\bigcup_{S \in \mathcal{S}'} S\right|$.
% In other words, the number of sets in each subcollection $\mathcal{S}'$ is less
% than or equal to the number of distinct elements in the union over the
% subcollection $\mathcal{S}'$.
% Hall's theorem states that $\mathcal{S}$ has a SDR if and only if
% $\mathcal{S}$ satisfies the marriage condition.

% \paragraph{Bipartite graphs.}

% A graph $G=(V,E)$ is \emph{bipartite} if 
% $V$ can be partitioned into two sets $U$ and $W$ such that each edge
% joins a vertex in $U$ to a vertex in $W$, and the pair $(U, W)$ is called a
% \emph{bipartition} of $V$ (or of $G$).
% The \emph{reduced adjacency matrix} $B = [b_{i,j}]$ is an $m \times n$
% $(0,1)$-matrix defined by $b_{i, j} = 1$ if and only if
% there is an edge beteween vertex $i$ of $U$ and vertex $j$ of $W$.
% A matching $M$
% in $G$ is a collection of edges no two of which have a vertex in
% common. If $v$ is a vertex and there is an edge of $M$ containing $v$,
% then $v$ \emph{meets} the matching $M$ and the matching $M$ meets the vertex
% $v$. A \emph{perfect matching} of $G$ is a matching
% that meets all vertices of $G$.

% \paragraph{Directed graphs.}

% A \emph{directed graph} (for short, a \emph{digraph}) $G$
% consists of a finite set $V$ of elements called vertices and a set $E$ of
% ordered pairs of vertices called arcs.
% The \emph{adjacency matrix} associated to $G$ is the $|V| \times |V|$
% $(0,1)$-matrix $A = [a_{i,j}]$ defined by $a_{i,j} = 1$ if and only if
% there is an arc from vertex $i$ to vertex $j$ in $G$.

%%
%% ---- Properties ----
%%
\section{Answering Wilf's question}%Hardness result}
\label{section:Hardness result}

We prove in this section that, given a square $(0,1)$-matrix $A$, deciding whether there 
exists a permutation matrix $P$ and a permutation matrix $Q$ of suitable size
such that $PAQ$ is triangular is \NPC.

\subsection{Disproving a previous related result}

Before giving our proof, it is worth mentioning that the following problem
(called LBQIS$(n,k)$ and rephrased to fit the context of this paper) is
claimed to be \NPC in~\cite{DasGuptaJiangKannanLiSweedyk:1998}:
Given a $(0,1)$-matrix of order $n$ and positive integer $k \leq n$,
do there exist permutation matrices $P$ and $Q$ such that
$
PAQ =
\begin{bmatrix}
A_{1,1} & A_{1,2} \\
A_{2,1} & A_{2,2}
\end{bmatrix}
$
with $A_{1,2}$ a square lower triangular matrix of size $k$ by $k$?
It is not very difficult to find a polynomial transformation from
LBQIS to Wilf's question, which would prove the \NPcompleteness of the
latter. Just add $n-k$ empty rows and $n-k$ empty columns 
to matrix $A$ to obtain a new matrix $A'$. Now, notice that each 
submatrix $A_{1,2}$ in a solution for LBQIS may be completed with the 
$n-k$ empty rows put before row 1 of $A_{1,2}$ and with the $n-k$ empty 
columns put after column $k$ of $A_{1,2}$ to yield a solution for the 
instance $A'$ in Wilf's question, and viceversa.

Unfortunately, paper~\cite{DasGuptaJiangKannanLiSweedyk:1998} 
contains a serious flaw in the proof. To fix things, note that
in~\cite{DasGuptaJiangKannanLiSweedyk:1998} LBQIS is stated in
terms of bipartite graphs, for which matrix $A$ is the reduced adjacency
matrix. Then, LBQIS$(n,k)$ is proved \NPC by reduction 
from another problem on bipartite graphs called LBIS$(n,k)$, using the so-called 
Rearrangement Lemma (Lemma 3.5 in ~\cite{DasGuptaJiangKannanLiSweedyk:1998}).
Two affirmations in the proof of this lemma are contradicted by the following
example. Let $G$ be the graph (input for LBIS) with vertices $U=\{i\, |\, 1\leq i\leq 4\}$
and $V=\{i\, |\, 1\leq i\leq 4\}$, whose edges are $(1,1), (2,1), (2,2),
(3,2), (3,4), (4,3)$ and $(4,4)$. Thus, $n=4$. Define $k=1$.  
Let $G'$ be the input graph for LBQIS built as in~\cite{DasGuptaJiangKannanLiSweedyk:1998},
and $k'=k^2+k=2$. Then the vertex subset $U'\cup V'$ of $G'$, with $U'=\{[2,4], [1,2]\}$
and $V'=\{[1,3], [2,2]\}$ is a solution of LBQIS of size $k'$ for which the
assumption on the first line of the Rearrangement Lemma's proof is false.
Also, the vertex subset $U'\cup V'$ of $G'$, with $U'=\{[1,1], [2,1]\}$
and $V'=\{[1,2], [1,3]\}$ is a solution of LBQIS of size $k'$ for which
the second affirmation in the same lemma (``clearly $q_1\leq p_1$'') is also false.

\subsection{Our \NPcompleteness proof for Wilf's question}

We present our results in terms of \SBL\ for configurations
of subsets. The rationale for considering \SBL\ for configurations
of subsets stems from the following lemma.

\begin{lemma}
    \label{lemma:bounded stepwise property <=> PAQ triangular}
    Let $\mathcal{S} = (S_i : 1 \leq i \leq n)$ be a configuration of subsets of
    some ground $n$-set, and let $A$ be the corresponding incidence matrix.
    There exist permutation matrices $P$ and $Q$ of order $n$
    such that $PAQ \leq \lltm_{\,n}$ if and only if
    there exists an \SBL\ of $\mathcal{S}$.
\end{lemma}

We need to focus our attention on a special type of \SBL. Call a bijective mapping
$\varphi : \mathcal{S} \to [n]$ \emph{normalized} if $\varphi$
maps the identical subsets of elements of $\mathcal{S}$
to a set of consecutive integers.
Most of the interest in normalized bijective labelings
stems from the following intuitive lemma.

\begin{lemma}
  \label{lemma:normalized stepwise bounded}
  Let $\mathcal{S} = (S_i : 1 \leq i \leq n)$ be a configuration of subsets of
    some ground $n$-set.
    If there exists an \SBL\ of $\mathcal{S}$
    then there exists a normalized \SBL\ of $\mathcal{S}$.
\end{lemma}

We are now ready to prove that
deciding whether there exists an \SBL\ of some
configuration of subsets is \NPC thereby proving that 
deciding whether a square $(0,1)$-matrix is a \PET\ matrix is \NPcomplete as well.
The proof proceeds by a reduction from the \PB{3Sat} problem - which is a known
\NPcomplete problem \cite{Cook:1971}.
% (see also
% \cite{Garey:Johnson:1979} and \cite{Papadimitriou:1994} for variants).
Let an arbitrary
instance of the \PB{3Sat} problem be given by a $3$CNF formula 
$\phi = c_1 \vee c_2 \vee \ldots \vee c_m$ over variables 
$x_1, x_2, \ldots, x_n$. 
Our construction is divided into two steps: 
(1) construction of a (polynomial size) ground set $\mathbf{X}$ and 
(2) construction of a configuration of subsets $C$ of the ground set
$\mathbf{X}$. 
Throughout the proof, parts of the ground set
$\mathbf{X}$ are written as capital bold letters 
($\mathbf{V}, \mathbf{T}, \mathbf{F}, \ldots$)
and subsets of the configuration are written with capital calligraphic
letters ($\mathcal{V}_i, \mathcal{T}_i, \mathcal{F}_i, \ldots$).

  \begin{figure}[H]
    \centering
    \begin{minipage}[c]{\textwidth}
      \centering
      \input{r.pstex_t}
      \caption{\label{fig:stepwise bounded labeling construction}
          Illustration of the construction 
          % used in the proof of 
          % Proposition~\ref{proposition:Stepwise Bounded Labeling is NPC} 
          for the $3$CNF formula 
          $\phi = (x_1 \vee x_2 \vee x_3) \wedge 
          (\overline{x}_1 \vee \overline{x}_2 \vee x_3) \wedge 
          (\overline{x}_1 \vee x_2 \vee \overline{x}_3)$. 
          Identical subsets are not distinguishable
          in our representation. A satisfying truth assignment is
          given by $f(x_1) = \mbox{TRUE}$, $f(x_2) = \mbox{FALSE}$ and
          $f(x_3) = \mbox{FALSE}$. For sake of clarity, neither
          the ground set $\mathbf{X}$ nor the collection of subsets
          $C$ is fully represented. 
          % Note that we have here dropped the
          % assumption that each clause is satisfied by its first
          % literal
          } 
    \end{minipage}%
  \end{figure}

To begin with, define 
$p_i = 3(n + m + 1 - i) + 2$,
$q_i = 3(n + m + 1 - i) + 1$ and
$r_i = 3(n + m + 1 - i)$ for $1 \leq i \leq n + m$.
Furthermore, define $p_{n+m+1}=1$,
$K = \sum_{i=1}^{n} q_i + 2 \sum_{i=n+1}^{n+m} q_i$ and
$L = \sum_{i=1}^{n+m} (p_{i+1} + r_i)$.
Let us now define the ground set $\mathbf{X}$. 
Consider the pairwise disjoint sets defined as follows:
$\mathbf{V}_i = \{v_{i, j} \mid 1 \leq j \leq p_i\}$,
$\mathbf{V}'_i  = \{v'_{i, j} \mid 1 \leq j \leq r_i\}$,
$\mathbf{T}_i = \{t_{i, j} \mid 1 \leq j \leq q_i\}$,
$\mathbf{F}_i = \{f_{i, j} \mid 1 \leq j \leq q_i\}$
for $1 \leq i \leq n$. Furthermore, define
$\mathbf{C}_i = \{c_{i, j} \mid 1 \leq j \leq p_{n + i}\}$
$\mathbf{C}'_i = \{c'_{i, j} \mid 1 \leq j \leq r_{n + i}\}$
for $1 \leq i \leq m$, and
$\mathbf{L}_{i, k} = \{\ell_{i, k, j} \mid 1 \leq j \leq q_{n + i}\}$
for $1 \leq i \leq m$ and $1 \leq k \leq 3$.
Finally, define $\mathbf{S} = \{s\}$.
For simplicity of notation, write
$\mathbf{V} = \bigcup_{1 \leq i \leq n} \mathbf{V}_i$,
$\mathbf{V}'= \bigcup_{1 \leq i \leq n} \mathbf{V}'_i$,
$\mathbf{T} = \bigcup_{1 \leq i \leq n} \mathbf{T}_i$,
$\mathbf{F} = \bigcup_{1 \leq i \leq n} \mathbf{F}_i$,
$\mathbf{C} = \bigcup_{1 \leq i \leq m} \mathbf{C}_i$,
$\mathbf{C}' = \bigcup_{1 \leq i \leq m} \mathbf{C}'_i$,
and
$\mathbf{L}_i = \bigcup_{1 \leq k \leq 3} \mathbf{L}_{i, k}$
for $1 \leq i \leq m$
and
$\mathbf{L} = \bigcup_{1 \leq i \leq m} \mathbf{L}_i$.
Informally, elements of $\mathbf{V} \cup \mathbf{V}'$ are associated to variables, 
elements of $\mathbf{T} \cup \mathbf{F}$ are associated to literals, elements of 
$\mathbf{C} \cup \mathbf{C}'$ are associated to clauses, elements of $\mathbf{L}$ 
are associated to literals in clauses and $\mathbf{S}$ is a  
\emph{separator set}. The ground set $\mathbf{X}$ of our construction is defined to be
$\mathbf{X} = 
\mathbf{V} \cup \mathbf{V}' \cup \mathbf{T} \cup 
\mathbf{F} \cup \mathbf{C} \cup \mathbf{C}' \cup 
\mathbf{L} \cup \mathbf{S}$.

Having defined the ground set $\mathbf{X}$, 
we now turn to the detailed construction of a configuration of subsets $C$ of $\mathbf{X}$. 
For sake of clarity, this will be divided into several steps. 
First, each variable $x_i$, $1 \leq i \leq n$, is associated to identical subsets 
$\mathcal{V}_{i, j}$, $1 \leq j \leq q_i$, in $C$. These subsets are defined as follows:
$\mathcal{V}_{i, j} = 
\left( \bigcup_{1 \leq k \leq i} \mathbf{V}_k \right) \cup 
\left( \bigcup_{1 \leq k \leq i-1} \mathbf{V}'_k \right)$
for $1 \leq i \leq n$ and $1 \leq j \leq q_i$.
Let us denote by $\mathcal{V}_i$, $1 \leq i \leq n$, 
the collection $(\mathcal{V}_{i, j} \mid 1 \leq j \leq q_i)$.
Next, each (positive) literal $x_i$, $1 \leq i \leq n$, 
is associated to identical subsets $\mathcal{T}_{i, j}$, $1 \leq j \leq r_i$, 
and to identical subset $\mathcal{T}'_{i, j}$, $1 \leq j \leq p_{i + 1}$. 
These subsets are defined as follows:
$\mathcal{T}_{i, j} = \mathbf{T}_i \cup 
\left( \bigcup_{1 \leq k \leq i} \mathbf{V}_k \right) \cup 
\left( \bigcup_{1 \leq k \leq i-1} \mathbf{V}'_k \right)$
for $1 \leq i \leq n$ and $1 \leq j \leq r_i$, and
$\mathcal{T}'_{i, j} = \mathbf{T}_i \cup \left( 
\bigcup_{1 \leq k \leq i} \mathbf{V}_k \right) \cup 
\left( \bigcup_{1 \leq k \leq i} \mathbf{V}'_k \right)$
for $1 \leq i \leq n$ and $1 \leq j \leq p_{i+1}$.
Of course, a similar construction of subsets applies for the negation 
$\overline{x}_i$ of each variable $x_i$, \emph{i.e.},
$\mathcal{F}_{i, j} = 
\mathbf{F}_i \cup \left( \bigcup_{1 \leq k \leq i} 
\mathbf{V}_k \right) \cup \left( \bigcup_{1 \leq k \leq i-1} 
\mathbf{V}'_k \right)$ for $1 \leq i \leq n$ and $1 \leq j \leq r_i$, and
$\mathcal{F}'_{i, j} = \mathbf{F}_i \cup \left( 
\bigcup_{1 \leq k \leq i} \mathbf{V}_k \right) \cup 
\left( \bigcup_{1 \leq k \leq i} \mathbf{V}'_k \right)$ 
for $1 \leq i \leq n$ and $1 \leq j \leq p_{i+1}$.
For readability, write 
$\mathcal{T}_i = (\mathcal{T}_{i, j} \mid 1 \leq j \leq r_i)$, 
$\mathcal{T}'_i = (\mathcal{T}'_{i, j} \mid 1 \leq j \leq p_{i+1})$, 
$\mathcal{F}_i = (\mathcal{F}_{i, j} \mid 1 \leq j \leq r_i)$ and 
$\mathcal{F}'_i = (\mathcal{F}'_{i, j} \mid 1 \leq j \leq p_{i+1})$ for $1 \leq i \leq n$. 
Note that the following (strict) inclusions hold for all $1 \leq i \leq n$, 
$1 \leq j_1 \leq q_i$, $1 \leq j_2 \leq r_i$ and $1 \leq j_3 \leq p_{i+1}$: 
(i) $\mathcal{V}_{i, j_1} \subset \mathcal{T}_{i, j_2} \subset \mathcal{T}'_{i, j_3}$ and 
(ii) $\mathcal{V}_{i, j_1} \subset \mathcal{F}_{i, j_2} \subset \mathcal{F}'_{i, j_3}$.
We now turn to the $m$ clauses of the $3$CNF formula. 
Each clause $c_i$, $1 \leq i \leq m$, is associated to identical subsets 
$\mathcal{C}_{i, j}$, $1 \leq j \leq q_{n+i}$. These subsets are defined as follows:
$\mathcal{C}_{i, j} = \mathbf{V} \cup \mathbf{V}' \cup 
\left( \bigcup_{1 \leq k \leq i} \mathbf{C}_k \right) 
\cup \left( \bigcup_{1 \leq k \leq i-1} \mathbf{C}'_k \right)$
for $1 \leq i \leq m$ and $1 \leq j \leq q_{n+i}$.
Let us denote by $\mathcal{C}_i$, $1 \leq i \leq m$, 
the collection $(\mathcal{C}_{i, j} \mid 1 \leq j \leq q_{n+i})$. 
It is easily seen that $\mathcal{V}_{i, j_1} \subset \mathcal{C}_{k, j_2}$ 
for all $1 \leq i \leq n$, $1 \leq j_1 \leq q_i$, $1 \leq k \leq m$ and 
$1 \leq j_2 \leq q_{n+k}$.

Now, we consider the only part of the construction that depends on which 
literal occurs in which clauses. Denote by $\lambda_{i, k}$ the $k$-th literal 
of clause $c_i$, 
that is write $c_i = \lambda_{i, 1} \vee \lambda_{i, 2} \vee \lambda_{i, 3}$ 
for $1 \leq i \leq m$,
where each $\lambda_{i, k}$ is a variable or its negation. 
The $k$-th literal, $1 \leq k \leq 3$, of each clause $c_i$, $1 \leq i \leq m$, 
is associated to identical subsets $\mathcal{L}_{i, k, j}$, $1 \leq j \leq r_{n+i}$, 
and to identical subsets $\mathcal{L}'_{i, k, j}$, $1 \leq j \leq p_{n+i+1}$. 
These subsets are defined as follows:
$\mathcal{L}_{i, k, j} = \mathbf{V} \cup \mathbf{V}' \cup 
\mathbf{A}_k \cup \mathbf{L}_{i, k} \cup 
\left( \bigcup_{1 \leq \ell \leq i} \mathbf{C}_\ell \right) \cup 
\left( \bigcup_{1 \leq \ell \leq i-1} \mathbf{C}'_\ell \right)$
for $1 \leq i \leq m$, $1 \leq j \leq r_{n+i}$ and $1 \leq k \leq 3$ and
$\mathcal{L}'_{i, k, j} = \mathbf{V} \cup \mathbf{V}' \cup \mathbf{A}_k 
\cup \mathbf{L}_{i, k} \cup \left( \bigcup_{1 \leq \ell \leq i} \mathbf{C}_\ell \right) 
\cup \left( \bigcup_{1 \leq \ell \leq i} \mathbf{C}'_\ell \right)$ 
for $1 \leq i \leq m$, $1 \leq j \leq p_{n+i+1}$ and $1 \leq k \leq 3$, 
where $\mathbf{A}_k = \mathbf{T}_\ell$ if $\lambda_{i, k} = x_\ell$ 
and $\mathbf{A}_k = \mathbf{F}_\ell$ if $\lambda_{i, k} = \overline{x}_\ell$. 
For the sake of clarity, write 
$\mathcal{L}_{i, k} = (\mathcal{L}_{i, k, j} \mid 1 \leq j \leq r_{n+i})$ and 
$\mathcal{L}'_{i, k} = (\mathcal{L}'_{i, k, j} \mid 1 \leq j \leq p_{n+i+1})$ 
for $1 \leq i \leq m$ and $1 \leq k \leq 3$. 
Again, observe that 
$\mathcal{C}_{i, j_1} \subset \mathcal{L}_{i, k, j_2} \subset \mathcal{L}_{i, k, j_2}$ 
for all $1 \leq i \leq m$, $1 \leq j_1 \leq q_{n+i}$, $1 \leq j_2 \leq r_{n+i}$, 
$1 \leq j_3 \leq p_{n+i+1}$ and $1 \leq k \leq 3$.

Our construction ends with $p_1 + K - 1$ \emph{utility subsets}. 
These subsets will be partitioned into two separate classes according to their 
indented function: \emph{bootstrap subsets} and \emph{separator subsets}. 
First, $C$ contains identical bootstrap subsets 
$\mathcal{B}_i$, $1 \leq i \leq p_1 - 1$, defined as follows:
$\mathcal{B}_i = \emptyset$ for $1 \leq i \leq p_1 - 1$. The idea is to force any \SBL\ to map the $p_1 - 1$ 
empty sets of $\mathcal{B}$ to the first $p_1  - 1 = 3(n + m) + 1$ integers. 
Indeed, it is easily seen that all the above defined subsets of the configuration 
of subsets $C$ but those of $\mathcal{B}$ contain at least $p_1$ elements and 
hence cannot be mapped to an integer $i \leq p_1 - 1$ in any \SBL\ of 
$C$. 
Second, $C$ contains identical separator subsets $\mathcal{S}_i$, $1 \leq i \leq K$, 
defined by:
$\mathcal{S}_i = 
\mathbf{V} \cup \mathbf{V}' \cup \mathbf{C} \cup \mathbf{C}' \cup \mathbf{S}$
for $1 \leq i \leq K$.
The rationale of these subsets is that we need a separator between subsets in $C$ 
corresponding to a satisfying truth assignment $f$ for the $3$CNF formula $\phi$ 
and garbage subsets of $C$, that is subsets not involved in the satisfying truth 
assignment $f$. For simplicity, let us denote by $\mathcal{B}$ the collection 
$(\mathcal{B}_i \mid 1 \leq i \leq p_1 - 1)$ and by $\mathcal{S}$ 
the collection $(\mathcal{S}_i \mid 1 \leq i \leq K)$. 
Clearly our construction can be carried on in polynomial time: indeed, we have
$|\mathbf{X}| = O(m^2 + n^2)$ and
$|C| = O(m^2 + n^2)$.

\begin{lemma}
  \label{lemma:Stepwise Bounded Labeling is NPC}
  There exists a satisfying truth assignment $f$ for $\phi$ if and only if 
  there exists an \SBL\ of the configuration of subsets 
  $C$ of the ground set $\mathbf{X}$.
\end{lemma}

>>>>

\usepackage[utf8x]{inputenc}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
%\usepackage{llncsdoc}
\usepackage[english]{babel}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{algorithmic}
\usepackage{float}
\usepackage{todonotes}
\usepackage{subfigure}

\usepackage{tikz}
\usetikzlibrary{matrix, calc, arrows}

\input{complexity}

@@

@@

\title{%
Unshuffling Permutations}%

\author{%
  Samuele Giraudo\inst{1} \and
  St\'ephane Vialette\inst{1}
}% end author
\institute{%
  Universit\'e Paris-Est, LIGM (UMR 8049), CNRS, UPEM, ESIEE Paris, ENPC, 
  F-77454, Marne-la-Vallée, France\\
  \email{vialette@univ-mlv.fr}
}% end institute
\date{\today}

@@

\begin{abstract}
bla bla bla \ldots
\end{abstract}

@@

@@

%%
%% ---- Notations ----
%%
\section{Notations}
\label{section:Notations}
